Exploring mathematical objects from custom-tailored mathematical universes

### 2018-05-25 slides-filmat2018.tex
Bonus slide on the little Zariski topos

\begin{frame}{First steps in alternate universes}
  \begin{changemargin}{-1.1em}{0em}
    \fontsize{10pt}{12pt}\selectfont
    \begin{itemize}
      \item \intex{
        $\Eff \models \text{``Any number is prime or is not prime.''}$
      }{\textcolor{green!90}{\cmark}}{
        There is a \hil{Turing machine} which determines of
        any given number whether it is prime or not.
      }

      \item \intex{
        $\Eff \models \text{``There are infinitely many prime numbers.''}$
      }{\textcolor{green!90}{\cmark}}{
        There is a \hil{Turing machine} producing arbitrarily many prime
        numbers.
      }

      \item \intex{
        $\Eff \models \text{``Any function~$\NN \to \NN$ is the zero function or not.''}$
      }{\textcolor{red!80}{\xmark}}{
        There is a \hil{Turing machine} which, given a Turing
        machine computing a function~$f : \NN \to \NN$, determines whether~$f$
        is the zero function or not.
      }

@@

      \item \intex{
        $\Sh(X) \models \text{``Any cont. function with opposite signs has
        a zero.''}$
      }{\textcolor{red!80}{\xmark}}{
        Zeros can locally be picked \hil{continuously} in
        continuous families of continuous functions.
        \textcolor{red}{(\href{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{video} for counterexample)}
      }
    \end{itemize}
  \end{changemargin}
\end{frame}

@@

  \hil{For example:}\ \ $\ZZ$,\ \ $\ZZ[X]$,\ \ $\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$,\ \ $\QQ$,\ \ $\RR$

@@

@@

\begin{frame}{Topos theory \ldots}
  \begin{itemize}
    \item enriches the platonism debate,
    %\item yields a new view on ``continuous dependence on parameters'',
    \item uncovers further relations between objects,
    \item allows to view objects from a different point of view,
    \item has applications in mathematical practice.
  \end{itemize}

>>>>

\begin{frame}{First steps in alternate universes}
  \begin{changemargin}{-1.1em}{0em}
    \fontsize{10pt}{12pt}\selectfont
    \begin{itemize}
      \item \intex{
        $\Eff \models \text{``Any number is prime or is not prime.''}$
      }{\textcolor{green!90}{\cmark}}{
        Meaning: There is a \hil{Turing machine} which determines of
        any given number whether it is prime or not.
      }

      \item \intex{
        $\Eff \models \text{``There are infinitely many prime numbers.''}$
      }{\textcolor{green!90}{\cmark}}{
        Meaning: There is a \hil{Turing machine} producing arbitrarily many
        primes.\\[0em]
      }

      \item \intex{
        $\Eff \models \text{``Any function~$\NN \to \NN$ is the zero function or not.''}$
      }{\textcolor{red!80}{\xmark}}{
        Meaning: There is a \hil{Turing machine} which, given a Turing
        machine computing a function~$f : \NN \to \NN$, determines whether~$f$
        is zero or not.
      }

@@

      \item \intex{
        $\Sh(X) \models \text{``Any cont. function with opposite signs has
        a zero.''}$
      }{\textcolor{red!80}{\xmark}}{
        Meaning: Zeros can locally be picked \hil{continuously} in
        continuous families of continuous functions.
        \textcolor{red}{(\href{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{video} for counterexample)}
      }
    \end{itemize}
  \end{changemargin}
\end{frame}

@@

  \hil{For instance:}\ \ $\ZZ$,\ \ $\ZZ[X]$,\ \ $\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$,\ \ $\QQ$,\ \ $\RR$

@@

{\renewcommand{\insertframenumber}{4a}
\begin{frame}{The little Zariski topos in more detail}
  Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.

  \begin{itemize}
    \item $\Sh(\Spec(A)) \models \text{``For all~$x \in A^\sim$, \ldots''}$

          Meaning: For all~$f \in A$ and all~$x \in A[f^{-1}]$, \ldots
          \medskip

    \item $\Sh(\Spec(A)) \models \text{``There is~$x \in A^\sim$ such that \ldots''}$

          \mbox{Meaning: There is a partition of unity,~$1 = f_1 + \cdots + f_n \in A$,}
          such that for each~$i$, there exists~$x_i \in A[f_i^{-1}]$
          with~\ldots
          \medskip

    \item $\Sh(\Spec(A)) \models \text{``$\varphi$ implies $\psi$''}$

          Meaning: For all~$f \in A$, if~$\varphi$ on stage~$f$, then~$\psi$ on
          stage~$f$.
  \end{itemize}
\end{frame}
\addtocounter{framenumber}{-1}}

@@

\begin{frame}{Topos theory \ldots}
  \vspace*{-0.5em}
  \begin{itemize}
    \item enriches the platonism debate,
    %\item yields a new view on ``continuous dependence on parameters'',
    \item uncovers further relations between objects,
    \item allows to view objects from a different point of view,
    \item has applications in mathematical practice.
  \end{itemize}
